A Method for Regularization of Evolutionary Polynomial Regressions

### 2013-12-20 texts/submission_02-asc/paper_spfga-review.tex
Completed Text revision upto subsection 2.2.

This representation of polynomials makes the problem of structure search very clear: except for the trivial cases, the number of possible monomials given $n$ variables and a maximum joint degree $d$ grows exponentially with either $n$ or $d$. But more importantly, by separating the set of monomials from the coefficients, the polynomial regression problem can be naturally split into two subproblems:
%
\begin{enumerate}
\item For a given set of monomials $\mathcal{Q} = \left\lbrace q_1, \ldots, q_k\right\rbrace$ find the regression coefficients $\Theta = \left\lbrace \theta_1,\ldots,\theta_k \right\rbrace$ that minimize the \ac{rmse} on a given dataset;

@@

%
This is a simple linear regression problem obtained by expanding $\mathcal{D}$ with columns that replicate the monomials in $\mathcal{Q}$. The second problem is treated in the \ac{GA} setting: Let $\mathcal{D}$ be a dataset as above and $\mathcal{P}$ a set of polynomials. For each polynomial $p\in \mathcal{P}$ let $\mathcal{Q}_p$ be the set of monomials in $p$ (without the coefficients) and define the (anti) fitness
%
\begin{eqnarray*}
\phi_p &=& \min_\Theta J_{\textrm{fit}}\left(\Theta;\mathcal{Q}_p,\mathcal{D}\right)
\end{eqnarray*}
%
by solving the first problem. With a fitness of every instance, the \ac{GA} genetic operators (usually mutation and crossover) evolve the population $\mathcal{P}$ until a reasonable approximation of a local minimum is found. 
%
\begin{algorithm}[t]
\begin{algorithmic}
\Function{EPR}{$D,pop_0,\epsilon, maxiter$}
	\State $pop \gets pop_0;\: err \gets 1.0+\epsilon$
	\While{$err > \epsilon \land iterations < maxiter$}
		\State $pop \gets \Call{IterateGA}{pop}$
		\State $pop \gets \Call{Sort}{pop, key = J}$\Comment{Sort population by regression error}
		\State $err \gets J\left(\Call{First}{pop}\right)$
	\EndWhile
	\State\Return $\Call{First}{pop}$
\EndFunction
\end{algorithmic}
\caption{This \ac{EPR} algorithm uses linear regression in the calculation of the \ac{rmse} $J$ and the space of polynomials is searched by the \acp{GA} iteration step. At exit the \ac{rmse} of the fittest instance is bounded by $\epsilon$ or the maximum number of allowed iterations.}\label{alg:gapoly}
\end{algorithm} 
%
Notice that the properties of \acp{GA} and linear regression entail that Algorithm \ref{alg:gapoly} converges to a polynomial that is a local minimum of the fitness function, encapsulated in the \ac{rmse} function $J_{\textrm{fit}}$.
%

@@

Let  $\left\lbrace q_1, \ldots, q_k\right\rbrace$ be a set of monomials over the variables $X_1, \ldots, X_m$. The encoding of that set using $d$ bits per exponent is a binary list such that
\begin{enumerate}
\item the initial segment of $k$ bits defines the active state of each monomial;
\item the remaining bits are split into $k$ segments of size $m\times d$, each representing a monomial;
\item the bits in each monomial segment are split into $m$ sub-segments of size $d$. The $j\nth$ sub-segment is the binary representation of the degree of the variable $X_j$ in the enclosing monomial segment;
\end{enumerate}
This encoding can also be viewed as the flattening of the binary exponents in the matrix representation prefixed by the activation segment. The set $\left\lbrace x_1^3 x_3, x_3^7, x_1 x_2\right\rbrace$ (then $m = 3$) has matrix representation
$$\left[ 
\begin{array}{ccc}
3 & 0 & 1 \\
0 & 0 & 7 \\
1 & 1 & 0
\end{array}
\right] = \left[ 
\begin{array}{ccc}
011 & 000 & 001 \\
000 & 000 & 111 \\
001 & 001 & 000
\end{array}
\right]_{\left( 2 \right)}
$$
where the right matrix is in binary form with  $d = 3$ bits. An encoding of this set of monomials with the extra monomial $x_1^6x_2^2x_3^5$ inactive, setting $k = 4$, would be
\begin{center}
1110;~011, 000, 001;~000,000,111;~001,001,000;~110,010,101
\end{center}
where, for reading purposes, semicolons separate segments and commas separate variables. The first $k=4$ bits inform that the first, second and third monomials are active while the fourth is not.

@@

The polynomial regression \ac{rmse} considered so far accounts for the ability to predict the transformed testset. A known problem of using a cost function based only in the dataset \ac{rmse} (and of polynomial regressions in general) is a tendency to overfit the training data. This excessive variance can be reduced by regularizing the \ac{rmse} function and one possible method is the inclusion of a penalty factor proportional to the number of monomials. Thus, the error from equation \ref{eq:rmse} defines the regularized version
\begin{eqnarray}
J_{\textrm{reg}}\left(\Theta;\mathcal{Q},\mathcal{D}\right) &=& \lambda^{k} J_{\textrm{fit}}\left(\Theta;\mathcal{Q},\mathcal{D}\right)\label{eq:rmse-reg}
\end{eqnarray}
%
where $k$ is the number of monomials in the polynomial. When $\lambda > 1$ polynomials with more monomials are penalized.

@@

\caption{Error distribution by regularization exponent for two common datasets. The box plots summarize error values of ten simulations for each value of $\lambda$. The smallest overall error, in grey, is achieved in both datasets when $\lambda=0.8$.}
\label{Abalone_dataset_lambdas}\label{Auto-MPG_dataset_lambdas}

@@

Figure~\ref{Abalone_dataset_lambdas} shows regularization results for the Abalone and Auto MPG datasets with ten runs for each $\lambda$ (the following section  includes information about these datasets). The typical inflection point lies around $\lambda = 0.8$. This somewhat paradoxical result -- since regularization tries to minimize the model's complexity in order to prevent overfitting -- is somewhat justified by the inclusion of the linear terms mentioned in the previous section. The linear terms `pushes' the model to linearity while a regularization term $\lambda < 1$ tries to add non-linearity terms. Within this tension, the algorithm is able to reduce the overall error, compared to the typical implementation \ac{EPR} found in the literature.

We'll denote by \ac{EPRR} the proposed extension of the EPR algorithm which includes the regularized cost function together with the new features in polynomial coding.

@@

\begin{quotation}
\note{Moved from previous section.} In the experiments that led to this article, the tests showed that the inclusion of the linear terms achieved a significant reduction of error rates. So, considering the previous example, the final set of monomials would be $\left\lbrace x_1^3 x_3, x_3^7, x_1 x_2, x_1, x_2, x_3 \right\rbrace$. The next section discusses this feature.
\end{quotation}

>>>>

This partial representation of polynomials makes the problem of structure search very clear: except for the trivial cases, the number of possible monomials given $n$ variables and a maximum joint degree $d$ grows exponentially with either $n$ or $d$. But more importantly, by separating the set of monomials from the coefficients, the polynomial regression problem can be naturally split into two subproblems:
%
\begin{enumerate}
\item For a given set of monomials $\mathcal{Q} = \left\lbrace q_1, \ldots, q_k\right\rbrace$ find the regression coefficients $\Theta = \left\lbrace \theta_1,\ldots,\theta_k \right\rbrace$ that minimize the \ac{rmse} on a given dataset;

@@

%
This is a simple linear regression problem obtained by expanding $\mathcal{D}$ with columns that replicate the monomials in $\mathcal{Q}$. The resulting dataset, $\mathcal{D} \cup \mathcal{Q}\left( \mathcal{D} \right)$, adds the monomial transformations in $\mathcal{Q}$ to the original dataset $\mathcal{D}$. An alternative formulation would just replace $\mathcal{D}$ by $\mathcal{Q}\left( \mathcal{D} \right)$. It turns out that the first formulation is a special case of the second (by including the variables in the monomial set) and has better \ac{rmse} performance --- what is not surprising because it uses more features.

The second problem is treated in the \ac{GA} setting: Let $\mathcal{D}$ be a dataset as above and $\mathcal{P}$ a set of polynomials. For each polynomial $p\in \mathcal{P}$ let $\mathcal{Q}_p$ be the set of monomials in $p$ (without the coefficients) and define the (anti) fitness
%
\begin{eqnarray*}
\phi_p &=& \min_\Theta J_{\textrm{fit}}\left(\Theta;\mathcal{Q}_p,\mathcal{D}\right)
\end{eqnarray*}
%
by solving the first problem. With a fitness of every instance, the \ac{GA} genetic operators (usually mutation and crossover) evolve the population $\mathcal{P}$ until a reasonable approximation of a local minimum is found. 
%
\begin{algorithm}[t]
\begin{algorithmic}
\Function{EPR}{$D,pop_0,\epsilon, maxiter$}
	\State $pop \gets pop_0;\: err \gets 1.0+\epsilon$
	\While{$err > \epsilon \land iterations < maxiter$}
		\State $pop \gets \Call{IterateGA}{pop}$
		\State $pop \gets \Call{Sort}{pop, key = J}$\Comment{Sort population by regression error}
		\State $err \gets J\left(\Call{First}{pop}\right)$
	\EndWhile
	\State\Return $\Call{First}{pop}$
\EndFunction
\end{algorithmic}
\caption{This \ac{EPR} algorithm uses linear regression for the calculation of the \ac{rmse} $J$ and the space of polynomials is searched in the \acp{GA} iteration step. At exit the \ac{rmse} of the fittest instance is bounded by $\epsilon$ or the maximum number of allowed iterations.}\label{alg:gapoly}
\end{algorithm} 
%
Notice that the properties of \acp{GA} and linear regression entail that Algorithm \ref{alg:gapoly} converges to a polynomial that is a local minimum of the fitness function, encapsulated in the \ac{rmse} function $J_{\textrm{fit}}$.
%

@@

Let  $\left\lbrace q_1, \ldots, q_k\right\rbrace$ be a set of monomials over the variables $X_1, \ldots, X_m$. The encoding of that set using $d$ bits per exponent is a binary list such that
\begin{enumerate}
\item the initial segment of $k$ bits defines the active state of each monomial;
\item the remaining bits are split into $k$ segments of size $m\times d$, each representing a monomial;
\item the bits in each monomial segment are split into $m$ sub-segments of size $d$. The $j\nth$ sub-segment is the binary representation of the degree of the variable $X_j$ in the enclosing monomial segment;
\end{enumerate}
This encoding can also be viewed as the flattening of the binary exponents in the matrix representation prefixed by the activation segment. The set $\left\lbrace x_1^3 x_3, x_3^7, x_1 x_2\right\rbrace$ (with $m = 3$) has matrix representation
$$\left[ 
\begin{array}{ccc}
3 & 0 & 1 \\
0 & 0 & 7 \\
1 & 1 & 0
\end{array}
\right] = \left[ 
\begin{array}{ccc}
011 & 000 & 001 \\
000 & 000 & 111 \\
001 & 001 & 000
\end{array}
\right]_{\left( 2 \right)}
$$
where the right matrix is in binary form using $d = 3$ bits. An encoding of this set of monomials with the extra monomial $x_1^6x_2^2x_3^5$ inactive, setting $k = 4$, would be
\begin{center}
1110;~011, 000, 001;~000,000,111;~001,001,000;~110,010,101
\end{center}
where, for reading purposes, semicolons separate segments and commas separate variables. The first $k=4$ bits inform that the first, second and third monomials are active while the fourth is not.

@@

The polynomial regression \ac{rmse} considered so far accounts for the ability to predict the transformed test-set. A known problem of using a cost function based only in the dataset \ac{rmse} (and of polynomial regressions in general) is the tendency to overfit  training data. This excessive variance can be reduced by regularizing the \ac{rmse} function and one possible method is the inclusion of a penalty factor proportional to the number of monomials. In this line the error from equation \ref{eq:rmse} defines a regularized version
\begin{eqnarray}
J_{\textrm{reg}}\left(\Theta, \lambda;\mathcal{Q},\mathcal{D}\right) &=& \lambda^{k} J_{\textrm{fit}}\left(\Theta;\mathcal{Q},\mathcal{D}\right)\label{eq:rmse-reg}
\end{eqnarray}
%
where $k$ is the number of monomials in the polynomial. When $\lambda > 1$ polynomials with more monomials are penalized. The regularized extension of \ac{EPR}, denoted by \ac{EPRR}, includes the regularized cost function.

@@

\caption{Error distribution by regularization exponent for two common datasets. The box plots summarize error values of ten simulations for each value of $\lambda$. The smallest overall error, in grey, is achieved in both datasets when $\lambda=0.8$. Performance of the non-regularized \ac{EPR} is plotted in the line $\lambda = 1$.}
\label{fig:lambda.error-distribution}\label{Abalone_dataset_lambdas}\label{Auto-MPG_dataset_lambdas}

@@

A simple exploration on the effect of the regularization parameter is depicted in Figure~\ref{fig:lambda.error-distribution} where it is possible to observe that the typical inflection point lies around $\lambda = 0.8$. This value, favoring ``larger'' polynomials is justified by the balance of the data's non-linearity and polynomial complexity: below $\lambda = 0.8$, even penalized, larger monomial sets achieve better \ac{rmse} performance than smaller ones while above that value the size of the monomial set is excessive. Within this tension the overall error results reduced when compared to the non-regularized \ac{EPR} version.

@@
